\chapter*{Acknowledgments}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}

%TODO thank fred and committee

% It is common courtesy to seek permission before listing a person in acknowledgments, so that should be done for all. TODO
Special thanks to Derrell Lipman, Farzeen Harunani, Michael Tissenbaum, Drew Ditto, Kim Douglas, Molly Laden, Akira Kamiya, Ben Shapiro, Josh Sheldon, Hal Abelson, Mark Guzdial, Sally Fincher, and Benjamin Xie for their help throughout this work.

Thank you to my system testers, listed in full in the Appendix.

%Thank you to my system testers: Mike Staub, William Sokolowsky, Jr., Ambarish Roy, Justin Richer, Charles Kaneb, Nicholas McKinnon, Christian Quale, Joshua O. Carlson, Russell Bernstein, Alistair Stead, Peter Galvin, Jessica Spink, Tim Baird, Barry Sherman, Emily Jackson, Reed Spool, Caroline Fonseca, Jessica West, Christopher Rogacz, Marybeth Moriarty, Matthew Sherman, and Andrew Chanler.

Special thanks to Ashely Hale, the undergraduate research assistant who did a huge amount of the manual data coding. The analysis would not have been possible without her.

Thank you to my committee--- Lyn helped me lay the groundwork years ago, and continues to drive me towards what has become our common mission. Jay helped steer the work towards its logical conclusion, and I am grateful for his enthusiasm and his eye for structure. Of course, I offer the deepest gratitude to my thesis supervisor and long-time advisor, Fred Martin. Fred first inspired me to pursue my casual interest in education, and taught me how to focus and hone that into a passion, mission, and career. Thank you, Fred. 

I would especially like to thank all my friends and family who have helped me stay strong. Thank you, Michelle Dubuke, my thesis buddy.

Finally, to my wife Stacy, thank you for your endless love and support. I could not have done this without you.

%TODO thank stacy, family, friends (Michelle, )

This material is based upon work supported by the National Science Foundation under Grants DUE-1225719 and DRL-1433592.\footnote{Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author and do not necessarily reflect the views of the National Science Foundation.}

\renewcommand{\thefootnote}{\arabic{footnote}}