\chapter{Author Biography}

Mark Sherman attended University of Massachusetts Lowell, earning a  Bachelor of Science degree in Computer \& Electrical Engineering, in 2008, demonstrating a capstone project in capacitive touchscreen technology. He continued his studies at UMass~Lowell, moving to the Computer Science department, and earned a Master of Science degree in Computer Science in 2010, with a research focus in computer science education. His masters thesis ``Exploration of Natural Design Tendencies of Novice Engineers'' identified testing and iteration patterns that may contribute to successful completion of engineering puzzles.

Mark is a proud member of the UMass Lowell Engaging Computing Group, where he has worked since 2008.

Mark also serves as an App Inventor Master Trainer with MIT, where he visits winning teams of a national design competition, trains students and teachers in App Inventor, programming, and project management to facilitate the teams production of a functional mobile application.
Projects mentored:
\begin{itemize}
\item HeadTek: Hardware and software solution for recording and processing head impact data with real-time data
reporting to mobile devices. Suitable for football, bike riding, and other impact-wary activities. Arduino, Firebase noSQL cloud database, sensors, real-time data, mobile visualization. Pine Crest Middle School, Fort Lauderdale, FL, 2016.
\item A Look Inside: App to explore inner workings of everyday devices. Provided clickable images that zoom into subcomponents with explanatory text. Utilized cloud SQL database to provide updatable list of devices, images, text, maps of click zones, traversal maps. FusionTables database, 3D modeling, custom data parser design. Upper St. Clair High School, Pittsburgh, PA, 2015.
\item SnapDocs: Utilize cloud-based optical character recognition from document scanner app, provide text representation of image. Cheney Middle School, Fargo, ND, 2014.
\item ChowChecker: Search for foods and display allergens present. Utilized online food databases and parsed ingredients lists. Hampstead Academy High School, Hampstead, NH, 2013.
\item Recycle Bin: Identified nearby recycling centers, helped community members form new recycling initiatives. STEM Center Middle School, Fargo, ND, 2013
\end{itemize}

\noindent Undergraduate courses at UMass Lowell: Organization of Programming Languages, Computing I, and Exploring the Internet.

\noindent Workshops developed and/or instructed: CAITE professional development for App Inventor; UMass Lowell Summer Bridge program; MassBay Community College Summer Bridge program; Beauty \& Joy of Computing professional development; Introduction to Mobile Robotics for Professors at Chitkara University, Chandigarh, India; undergraduate intensive in Mobile Robotics in Guangzou, China. 
